Likelihoods for a general class of ARGs under the SMC

### 2025-04-28 review-responses.tex

\begin{point}{\revref} % 24:
 evol biol $\to$ genetics?
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 27:
 ancestral processes $\to$ genetic processes?
\end{point}

\reply{
We would prefer to keep ancestral processes here as ARGs contain
information about, for example, population structure which is not
a genetic process, but something that we gain insights into
through genetics.
}

@@

\begin{point}{\revref} % 138:
 is unary in the second tree.
\end{point}

\reply{
Done.
}

\begin{point}{\revref} % 140:
 all lineages with recombination event
\end{point}

\reply{
We  prefer ``hit by a recombination event" as it's a little more concrete,
and in keeping with standard phrasing in the field.
}

### 2025-04-28 paper.tex

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the
details of recombination events, and heuristic methods
omit these details in different ways. \revpoint{2}{8}
The significant progress made in
scalability by these heuristic methods is in part \emph{because} they don't
attempt to precisely reconstruct recombination events.

@@

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards-in-time.
At any point in time, the state of the process is
the set of labelled lineages extant at time $t$,
$L(t) = \{X_j(t)\}$.
A lineage $X_j(t) = [x_{j}, y_{j})$ is labelled by
an integer $j$ and defines the
half-open genomic interval of its ancestral material.
% Ea
% where each lineage $X_j(t)$ is labeled by an integer $j$
% and represented by a
% ancestral material.
% Any lineage can thus be summarized as
% for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards-in-time, $t=0$ is today, and
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit
of genomic length and time:
so, if $T(t) = \sum_j |X_j(t)|$ is total length of ancestral material
carried by lineages at time $t$,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
given new, unique labels. We then remove lineage $j$
and add the two new lineages to the state $L(t)$.
This operation keeps the total amount of ancestral material unchanged.

>>>>

There is considerable uncertainty in ARG inference, and often a fundamental
lack of information in the sequence data to distinguish different
possibilities. Consider, for example, a case in which we have three
ancestral lineages with identical sequences. We know that there
must have been two coalescence events, but there is no mutational information to
distinguish their relative ordering. Sampling methods overcome this problem
by randomly choosing an ordering, with the uncertainty communicated by the order
changing in the different ARGs sampled from the posterior.
The other approach (used by methods such as tsinfer) is to communicate
the uncertainty structurally by means of a polytomy: we have no
information about the intermediate coalescent event, and so we omit it.
Similarly, there is often a fundamental lack of information about the
details of recombination events, and heuristic methods
omit these details in different ways. \revpoint{2}{8}
The significant progress made in
scalability by these heuristic methods is in part because they don't
attempt to precisely reconstruct recombination events.

@@

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards-in-time.
At any point in time, the state of the process is
the set of labelled lineages extant at time $t$,
$L(t) = \{X_j(t)\}$.
A lineage $X_j(t) = [x_{j}, y_{j})$ is labelled by
an integer $j$ and defines the
half-open genomic interval of its ancestral material.
% Ea
% where each lineage $X_j(t)$ is labeled by an integer $j$
% and represented by a
% ancestral material.
% Any lineage can thus be summarized as
% for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards-in-time, $t=0$ is today, and
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to the next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit
of genomic length and time:
so, if $T(t) = \sum_j |X_j(t)|$ is the total length of ancestral material
carried by lineages at time $t$,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
given new, unique labels. We then remove lineage $j$
and add the two new lineages to the state $L(t)$.
This operation keeps the total amount of ancestral material unchanged.
